---
title: "The Effect of Language Dominance on Anticipatory Nasalization"
shorttitle: "Anticipatory Nasalization"
author:
  - name: Robert Esposito
    corresponding: true
    email: rme70@rutgers.edu
    affiliations:
      name: "Rutgers University"
      department: Department of Spanish and Portuguese
      address: 15 Seminary Pl.
      city: New Brunswick
      region: NJ
      country: USA
      postal-code: 08901
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    # Example: This study was registered at X (Identifier Y).
    study-registration: ~
    # Acknowledge and cite data/materials to be shared.
    data-sharing: ~
    # Example: This article is based on data published in [Reference].
    # Example: This article is based on the dissertation completed by [citation].  
    related-report: ~
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
number-sections: false
abstract: "Languages differ in their inventory of nasal segments, with only about a quarter of the world's languages having contrastive nasal vowels in their phonological inventory. However, all languages, to some degree, include phonetically-motivated nasalized vowels due to coarticulation in (NASAL-)VOWEL(-NASAL) ((N)V(N)) sequences, where at least one nasal consonant precedes or procedes the vowel. Languages differ in the degree of coarticulatory nasalization. For example, Mexican Spanish has been described as having fairly little nasal coarticulation, whereas American English has a relatively high degree of nasal coarticulation to the point that vowel nasalization has been said to be phonologized. It has been proposed that bilinguals must acquire these language-specific timings to sound native-like. The Complex Adaptive System Principles (CASP) model suggests that the bilingual brain pushes for convergence, especially in cases where the bilingual is not balanced, such that English-like timing patterns will occur in Spanish or vice-versa. To put this theory to the test, the present study used the The Multilingual Hispanic Speech in California (MuHSiC) corpus to investigate coarticulatory nasalization. The metric A1-P0 was used to measure nasalization at 10 points within vowels in /VN/ sequences, and then generalized additive linear models (GAMs) were fit to the time series. Differences in coarticulatory nasalization patterns were found across languages, but language dominance did not predict nasalization patterns within languages. These findings align with previous research on the effects of dominance on coarticulatory nasalization in Spanish-English bilinguals, but they go against the preedictions of the CASP model."
keywords: [bilingualism, coarticulation, nasal]
floatsintext: true
bibliography: "./lit/bi_coart_nasal.bib"
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    documentmode: man
---
```{r}
#| label: setup
#| include: false

# Set document defaults
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  fig.retina = 2, 
  echo = F, 
  message = F, 
  warning = F,
  fig.asp = 0.5625,
  out.width = "100%",
  dpi = 300, 
  fig.path = 'figs/', 
  dev = c("png", "pdf")
  )

# load libs
source(here::here("scripts","r","06_manuscript_stats.R"))
```

Bilinguals have the task of acquiring two different phonological systems. Parts of a language's phonological system is non-contrastive, such as nasality for vowels in English and Spanish [@wals-18]. Although vowel nasality is non-contrastive in these languages, vowels become partially or fully nasalized when adjacent to a nasal consonant [@sampson1999nasal]. Monolingual varieties of (American) English and (Peninsular) Spanish have been found to differ in timing patterns for coarticulatory vowel nasalization, such that English is, on the whole, more nasal than Spanish [@sole1991differentiating]. These differences have been replicated in American English-Mexican Spanish bilinguals [@beristain2023airflow], but it is currently unclear if individual differences, such as language dominance, may impact the production (and, inferentially, the storage) of these language-specific timing patterns. The Complex Adaptive System Principles (CASP) model [@filipovic2019complex; @hawkins2024bilingualism] predicts that language dominance interacts with the maxim [maximize common ground]{.smallcaps}, such that balanced bilinguals will produce more monolingual-like patterns in each of their languages, whereas unbalanced bilinguals will favor forms in both languages that converge on forms in their dominant language. As far as the author knows, there has not yet been research explicitly testing the CASP model on phonological or phonetic data. The current study addresses this gap by investigating coarticulatory vowel nasalization in simultaneous and early sequential bilinguals in the Multilingual Hispanic Speech in California [MuHSiC; @amengual2026muhsic] corpus. A1-P0, a measure of nasalization that has previously been used on corpora [@chen1997acoustic; @tamminga2015cross], was extracted at 10 points within vowels in /VN/ sequences and used to fit non-linear models that addressed differences across and within languages. The present study combines sociophonetic, bilingual, and articulatory phonology domains to better understand what phonological information is stored in a bilingual's language systems.

## Heritage Speakers

Heritage speaker populations have received comparatively little attention with regards to coarticulatory phenomena. It has been established that heritage speakers tend to maintain phonemic robustness [@kissling2018exploratory], but investigation of fine-grained coarticulatory phenomena will provide richer detail to the monolingual and bilingual grammars. In the current case, whether or not gestural timing is encoded in the grammar, in the form of coarticulatory vowel nasalization, can be determined. @beristain2023airflow investigated 16 heritage Mexican Spanish L2 English second generation speakers of varying dominance from the Chicagoland area using a nasometer to determine if they present distinct time setting patterns for nasality in English and Spanish /VN/ structures. An effect for language was found, but not for language dominance, which was measured with the Bilingual Language Profile [BLP; @birdsong2012bilingual]. On the other hand, @amengual2018asymmetrical found an effect for dominance, again measured with the BLP, on the acoustic realization of laterals in 40 Spanish-English heritage speakers. 

## CASP Model

The Complex Adaptive System Principles (CASP) model [@filipovic2019complex; @hawkins2024bilingualism] proposes five principles that underlie language behavior in bilinguals:

(a) minimize learning effort
(b) minimize processing effort
(c) maximize expressive power
(d) maximize efficiency in communication
(e) maximize common ground

Coarticulatory vowel nasalization, although allophonic in English, does not contrast in meaning with oral vowels in English or Spanish. As such, bilinguals may (a) [minimize learning effort]{.smallcaps} by not encoding language-specific gestural timing information for nasals or (e) [maximize common ground]{.smallcaps} by sharing a single gestural timing representation for both languages. Given that previous research has demonstrated that bilinguals do maintain language-specific timing information [@beristain2023airflow], it is relevant that CASP also takes into account individual factors, such as language dominance, as it has also been found that gestural timing encoding varies with language dominance [@amengual2018asymmetrical]. However, such dominance effects have yet to be seen for coarticulatory vowel nasalization in English-Spanish bilinguals, most likely due to low sample size and little dominance variation in the sample [@beristain2023airflow].

The CASP model has little to say about phonological or phonetic interactions in the bilingual grammar, but the principles may be applied to such a situation as this. For example, at the level of form @filipovic2019complex claim that high-proficiency bilingual speakers better control their output than low-proficiency bilinguals and "know" when they can [maximize common ground]{.smallcaps}, such that they will not produce "ungrammatical" structures in either language. In this case, high degree of vowel nasalization in Spanish will be seen as divergent behavior, while low degree of vowel nasalization in English will be equally seen as divergent.

## Nasal Segments

98% of the world's languages have at least one nasal consonant [@wals-18], while about 26% percent of all languages have contrastive nasal vowels [@wals-10], such as Brazilian Portuguese or French. Spanish and English are among the languages that do not have contrastive nasal vowels, but vowels still become nasalized in vowel-nasal (VN) syllables due to timing gestures and biomechanical production of segments [@sampson1999nasal]. 'Nasalization' occurs when the velum lowers and velopharyngeal (VP) coupling occurs, such that the airstream flows through both the oro-pharyngeal and nasal tracts [@chafcouloff1999velopharyngeal; @barlaz2018oropharygneal]. When a vowel segment is followed by a nasal consonant, the velum anticipatorily lowers resulting in VP coupling and partial or full nasalization of the vowel [@zellou2022coarticulation]. @fig-gestural-overlap represents such a process in English (upper) and Spanish (lower).

```{r}
#| label: fig-gestural-overlap
#| echo: false
#| fig-cap: A representation of English (upper) and Spanish (lower) gestural timings of the word 'pan' (in Spanish, *bread*). Note that the nasal gesture overlaps the vowel gesture much more in English than in Spanish. Reproduced without permission from @goodin2016nasals.

knitr::include_graphics(
  here("figs", "gestural_overlap.jpg")
  )
```

Coarticulatory vowel nasalization varies cross-linguistically [@clumeck1976patterns]. @sole1991differentiating found that Peninsular Spanish and American English differ in degree of nasalization based on speaking rate, concluding that nasalization in the former is phonetic and in the latter, phonological. These findings align with more recent observations of English monolingual populations that nasalization is extensive in American English [@kramer2019vowel], whereas many varieties of Spanish show a much shorter gestural overlap [@bongiovanni2021acoustic].

## Measures of Nasalization

Degree of nasalization may be measured using a nasometer [@bressmann2021nasometry], but in its absence, a number of acoustic features have been proposed to correlate with nasalization such as reduction of formant amplitudes, widening of formant bandwidths, modulation of formant frequencies, shifting of spectral energy toward lower frequencies, and addition of poles (i.e., formants) and zeros (i.e., anti-formants) to the acoustic spectrum (@carignan2018using; @chen1997acoustic; @feng1996some; @styler2017acoustical; inter alia).

In the current study, nasalization will be measured using A1-P0 [@chen1997acoustic; @tamminga2015cross]. This measure is the difference between the amplitude A1 of the first formant and the amplitude P0 of the nasal peak at low frequencies, and it works particularly well for mid and low vowels. @carignan2021practical compared various metrics with nasometer data across various English vowels and found that A1-P0 had an overall correlation with the nasometer "ground-truth" data of 0.66. However, looking at individual vowels demonstrates that A1-P0 has a 0.76 correlation with the nasometer data for the vowel /a/. As such, in the absence of a nasometer, A1-P0 provides an approximate metric to determine nasalization for specific vowels.

## History of Spanish-English Contact in California

Today, California has the largest Hispanic population in the United States [@lopez2015california], characterized by long-standing Latino communities and newly arrived Latinos [@zentella1997growing]. While Spanish is being maintained at the societal level because of the influx of new immigrants, at the individual and family level, there is a clear shift to English [@silva1994language]. Residents of long-standing Latino communities include both newly arrived Latinos who are monolingual in Spanish, bilingual Latinos who have acquired English and retain Spanish, young bilingual Latinos who have acquired both languages in their communities, and monolingual English-speaking Latinos. In these communities, individuals interact with one another using Spanish, English, or both. Choice of language depends on language proficiency, topic of interaction, setting, and relationship between speakers. Children born and raised in such communities acquire language ability to interact with newly arrived individuals and monolingual members of the majority Anglophone community [@zentella1997growing]. Spanish is primarily restricted to low-level functions and private sphere interactions [@valdes2004learning].

# The Present Study

The present investigation presents an opportunity to relate articulatory phonology, sociolinguistics, and bilingualism. Sequences of vowel-nasal (VN) in American English and Mexican Spanish, specifically /an/ and /am/, are examined. Both Spanish and English possess /a/ in their phonological inventories, and A1-P0, as explained above, is an adequate measure of nasalization for this vowel, which will be used to identify degree of nasalization within the vowel. Spanish and English also differ in nasal consonant inventories, so the present investigation is limited to only those nasal consonants that are shared between the two phonological inventories. The variation investigated is degree of vowel nasalization, as measured with A1-P0. As such, the collected data was selected to answer the following research questions:

1) Do bilinguals produce language-specific timing patterns of nasal gestures in English and Spanish?

Previous research has demonstrated that bilinguals do produce language-specific timing information of gestures for nasals [@beristain2023airflow], which would be reflected in differing patterns of A1-P0 across the vowel in VN sequences. Specifically, more negative values of A1-P0, indicating greater degree of nasality, are expected in English, whereas more positive values are expected in Spanish. Such patterning would indicate that bilinguals store language-specific timing patterns in each of their languages.

2) Does language dominance predict degree of nasalization in each language in /aN/ sequences?

The CASP model [@filipovic2019complex; @hawkins2024bilingualism] predicts that language dominance interacts with the [maximize common ground]{.smallcaps} principle, such that balanced bilinguals will show greater maintenance of language-specific patterns than unbalanced bilinguals. It is hypothesized English-dominant bilinguals will show greater nasalization in Spanish compared to the average, whereas Spanish-dominant bilinguals will show less nasalization in English compared to the average. If it is found that unbalanced bilinguals produce timing patterns in line with their dominant language in their non-dominant language, such a finding would provide supporting evidence that language dominance interacts with the [maximize common ground]{.smallcaps} principle, even for differences that do not produce meaning contrasts.

## The MuHSiC Corpus

The Multilingual Hispanic Speech in California [MuHSiC; @amengual2026muhsic] corpus is an open-access collection of sociolinguistic interviews with bilinguals speakers of diverse social backgrounds and regional origins across California. It is comprised of 600 interviews, approximately 70 minutes per speaker, divided evenly between English and Spanish. It has over 42,000 minutes of transcribed and phonologically segmented speech, 4,698,607 time-aligned word tokens, and 16,277,689 time-aligned phonological segments. Included participants, who were primarily university-aged, had a minimum residency of five consecutive years in California and at least conversational proficiency in both English and Spanish. Each participant completed a socio-demographic questionnaire and the Bilingual Language Profile (BLP) questionnaire [@birdsong2012bilingual].

```{r}
#| label: fig-dominance
#| echo: false
#| fig-cap: Violin plot and boxplot of Bilingual Language Profile (BLP) scores in the sample. Possible BLP scores range from -218 (Spanish-dominant) to 218 (English-dominant). Jittered points represent individual BLP scores, color-graded based on dominance in relation to the possible score range.

knitr::include_graphics(
  here("figs", "p_dominance_distribution.png")
  )

```

This corpus provides high-quality audio that is ideal for acoustic analysis for the bilingual community under investigation [@amengual2026muhsic]. The interviewees in the corpus are primarily of Mexican origin (meaning that the interviwee, her father, mother, or at least one of four grandparents was born in Mexico). Of the `r mex_origin[1,2]` females in the corpus, `r mex_origin[1,1]` are of Mexican origin, or approximately `r mex_origin[1,3]`% of female interviwees. This fact is important, as Mexican varieties of Spanish tend to present less coarticulatory nasalization than other varieties, such as Caribbean varieties [@bongiovanni2021acoustic].

Participants from the corpus were filtered on specific socio-demographic information provided in the metadata. 1) Only interviewees who identified as `female`, which was selected to limit anatomical differences and to prevent interference of gender-related patterns of A1-P0 [@tamminga2015cross]. 2) The interviewee's parents both spoke English and/or Spanish and they did not speak another language to prevent any substantial input from another language, and thus possible influence on the English and Spanish language systems. 3) Related, the interviewee did not speak an additional language to English and Spanish to prevent any possible influence from another language system. Using these filters, `r num_interviewees` were identified for inclusion, each providing one interview in Spanish and one interview in English for a total of `r num_interviewees * 2` interviews. @fig-dominance shows distributional information of the interviewees' Bilingual Language Profile (BLP) scores.

## Token Identification and Extraction

Tokens for analysis were identified as contiguous segments /an/ and /am/ using the time-aligned TextGrids in the MuHSiC corpus [@amengual2026muhsic]. English transcriptions use ARPABET [@klautau2001arpabet], so tokens were extracted using the ARPABET symbol "AAN" and "AAM", which correspond to the IPA sequences /an/ and /am/. Despite other nasal consonants being available in English and Spanish, only /n/ and /m/ were selected as these overlap in their phonological inventories, whereas /ɲ/ is phonemic in Spanish but not in English, and /ŋ/ is phonemic in English but not in Spanish.

`NasalityAutomeasure.praat`[^1] [@styler2015acoustical] was used in `Full-Auto` mode to extract A1-P0 and vowel duration, as well as a number of other acoustic features associated with nasality that were not used for analysis here, at 10 equidistant points within each target vowel. The formant range was set to 5500 Hz since the interviewees were female [@carignan2021practical]. Since high vowels were not included, the highest and lowest "sane" F1 values were set to 1000 Hz and 500 Hz. The script was modified to document the segment immediately before and after the target vowel.

For the present analysis, only sequences of an unvoiced stops /p t k/, /a/, and a nasal consonant (henceforth represented as /TaN/) were included for analysis. This decision had multiple motivators. First, a previous nasal segment would potentially induce the vowel to already be nasalized, preventing any clear change in A1-P0 shape over time. Second, a preliminary analysis found that A1-P0 patterned distinctly in the first quarter of the vowel depending on the previous segment. As the present analysis is not interested in how the previous segment impacts A1-P0, the unvoiced stops were chosen due to (mostly) clear segment boundaries compared to, for example, fricatives or sonorants. Furthermore, the unvoiced stops form a more homogeneous class of sounds both within and across languages compared to, for example, voiced stops (which are often realized as approximants in Spanish) or sonorants (the segments of which differ between English and Spanish).

To clean the data, any tokens that were missing at least one of the ten measured values for whatever reason, such as the Praat script being unable to extract a value and returning `--undefined--`, were removed. Then, following @tamminga2015cross, the following data cleaning procedure was performed to account for probable measurement errors: 1) all vowels with a duration below 50 msec were removed. 2) A1-P0 values were z-scored for each language, and all values that exceeded $\pm$2.5 were removed, as well as all other values from their corresponding token. The decision to standardize A1-P0 within language is because it is assumed that the distribution of A1-P0 values are different for English and Spanish. @tbl-outliers provides information regarding outlier removal and number of tokens in the final dataset. A high proportion of data was removed from the dataset. It was deemed more imperative to remove possible measurement errors than to retain them due to the high number of tokens that remained.

```{r }
#| label: tbl-outliers
#| tbl-cap: Descriptive information on outliers for the English dataset, Spanish dataset, and the combined dataset.

ft_outliers <- flextable(outlier_summary) %>%
  colformat_double(j = c("n_values_with", "n_tokens_with",
                          "n_values_without", "n_tokens_without"),
                   digits = 0) %>%  
  bold(part = "header") %>%
    colformat_double(j = "proportion_retained", digits = 2) %>%
  autofit() %>%
  set_header_labels(
  language = "Language",
  n_values_with = "Original Values",
  n_tokens_with = "Original Tokens",
  n_values_without = "Trimmed Values",
  n_tokens_without = "Trimmed Tokens",
  proportion_retained = "Proportion Retained"
)

ft_outliers

```

[^1]: https://github.com/stylerw/styler_praat_scripts/tree/master/nasality_automeasure.

## Statistical Analysis

A1-P0 has previously been used to study coarticulatory nasalization in English corpora to find differences across generation, gender, and variety, but the statistical models built were linear [@tamminga2015cross]. The current study used a generalized additive mixed model (GAMMs) for statistical inference to model the non-linear effects of time on A1-P0 in each vowel. All GAMMs were fit using the `mgcv` package [@wood2017generalized; @wood2011fast] in R [@r2025].

Specifically, A1-P0 was the dependent variable, taken along the duration of Spanish and English /a/ vowels at 10 equidistant points from simultaneous and early-sequential bilinguals in the Multilingual Hispanic Speech in California [MuHSiC; @amengual2026muhsic] corpus. The final models reported here were compared with simpler models using the function `compareML()` [@itsadug2025r]. Only models that had significantly better fits than the ones they were compared to are reported here. Furthermore, all reported models are AR1 error models, which were fit due to high residual autocorrelation, typical of time series data [@soskuthy2017generalised]. All AR1 error models were compared to their corresponding non-AR1 model, again using `compareML()`[^2].

The first model reported addresses Research Question 1. Parametric terms for the variables `language` (English or Spanish) and `vowel duration` [@tamminga2015cross] were included to model mean differences in A1-P0. Smooth terms over `percent vowel` and `percent vowel` with the by-variable `language` were included to model non-linear trajectories of A1-P0 in each language. This model was set up so that the `percent variable by language` term represents the *difference* between Spanish and English, which can directly inform about significant differences between the shapes of the two trajectories, as opposed to if two smooths were fit for both languages [@soskuthy2017generalised].

Two similar models were fit to address Research Question 2. The first model reported was fit on the English data. Parametric terms for the variables `vowel duration` [@tamminga2015cross] and `Bilingual Language Profile (BLP) score` were included to model mean changes in A1-P0. Smooths terms over `percent vowel` and `percent vowel` with the by-variable `BLP score` were included to model non-linear trajectories of A1-P0 over time and by language dominance. The second model was fit similarly, except `vowel duration` not included as a parametric term because it was not found to be significant.

Statistical inference was completed through model comparisons using `compareML()`, p-values ($\alpha$ = 0.05) obtained from model outputs with the `summary()` function, and visual inspection of plots generated with `plot_diff()` and `plot_smooth()` [@itsadug2025r].

[^2]: More information on model comparisons can be found here: https://github.com/RobertEspo/bi_coart_nasal.

# Results

The following section reviews the models previously described. It is important to note that more negative values of A1-P0 indicate greater degree of nasalization.

## Across-Language Comparisons of A1-P0

@tbl-rq1-param and @tbl-rq1-smooth show the outputs from the model for parametric and smooth terms, respectively. The parametric difference term, `language.ordSpanish`, indicates that the overall mean for A1-P0 in Spanish is significantly higher than in English (represented by the intercept). Similarly, the difference smooth `s(point_vwlpct):language.ordSpanish` shows a significant difference, indicating a significance difference in shape of A1-P0 across normalized time in /a/ across English and Spanish.

```{r}
#| label: tbl-rq1-param
#| tbl-cap: Parametric terms for the model comparing A1-P0 across Spanish and English. English is the reference level. The difference term, `language.ordSpanish`, indicates that there is a significant difference in mean overall mean of A1-P0 across the two languages.

rq1_param_ft <- flextable(gam_rq1_ar1_param_df) %>%
  colformat_double(digits = 2) %>%
  bold(part = "header") %>%
  autofit()

rq1_param_ft

```

```{r}
#| label: tbl-rq1-smooth
#| tbl-cap: Smooth terms for the model comparing A1-P0 across Spanish and English. English is the reference level. The difference term, `s(point_vwlpct):language.ordSpanish`, indicates that there is a significant difference in trajectory of A1-P0 across the two languages.

rq1_smooth_ft <- flextable(gam_rq1_ar1_smooth_df) %>%
  colformat_double(digits = 2) %>%
  bold(part = "header") %>%
  autofit()

rq1_smooth_ft

```

@fig-rq1-smooths plots the smooths for fitted A1-P0 values in Spanish and English. Visual inspection indicates that Spanish displays overall less nasalization than English across the entirety of the vowel, but the estimates peak at different time points. Spanish A1-P0 peaks at `r round(peaks_smooth[2,2],2)`% of the vowel at `r round(peaks_smooth[2,3],2)` dB; English A1-P0 peaks at `r round(peaks_smooth[1,2],2)`% of the vowel at `r round(peaks_smooth[1,3],2)` dB. From those points onward, A1-P0 slopes downward. Visually, Spanish A1-P0 has a steeper slope than English, but still remains at an overall higher vowel to the end of the vowel.

@fig-rq1-diff plots the difference in A1-P0 values between Spanish and English, with English as the baseline (that is, Spanish values are subtracted from English). The trajectory does not intersect 0, indicating a significant difference in A1-P0 trajectory across English and Spanish. Spanish is overall more positive than English, with the greatest divergence occurring at `r round(valley_diff[1,1],2)`% of the vowel with an estimated difference in A1-P0 of `r round(valley_diff[1,1],2)` dB. Comparing @fig-rq1-smooths and @fig-rq1-diff indicates that as the end of the vowel is approached, Spanish A1-P0 lowers to approach English values, but remains significantly different from English as indicated by the confidence interval in @fig-rq1-diff not intersecting 0.

```{r}
#| label: fig-rq1-smooths
#| echo: false
#| fig-cap: Smooths of A1-P0 in English (blue) and Spanish (yellow) over normalized time in a vowel. The intersection of vertical and horizontal lines with the smooths indicate peak location for A1-P0. 

knitr::include_graphics(
  here("figs", "p_rq1_smooths.png")
  )

```

```{r}
#| label: fig-rq1-diff
#| echo: false
#| fig-cap: Smooths of A1-P0 in English (blue) and Spanish (yellow) over normalized time in a vowel. The intersection of the vertical and horizontal lines on the smooth indicate the greatest divergence point of A1-P0 between English and Spanish.

knitr::include_graphics(
  here("figs", "p_rq1_diff.png")
  )

```

## Effects of Language Dominance on A1-P0

The following two sections report the models used to test the effects of language dominance on A1-P0. First, the English-only model will be considered, and then the Spanish-only model. For plots, given the distribution of BLP scores in the sample (see @fig-dominance), predicted A1-P0 values for BLP scores of -75 (Spanish-dominant), 0 (balanced), and 75 (English-dominant) were used, although these values are fairly arbitrary. That is to say, the full range of BLP scores is -218 (Spanish-dominant) to 218 (English-dominant), but the sample had a score range of `r round((dominance_distribution$interviewee_final_blp_score %>% range())[1],2)` to `r round((dominance_distribution$interviewee_final_blp_score %>% range())[2],2)`, with a skew towards positive scores.

### English

The model output's parametric and smooth terms are displayed in @tbl-rq2-en-param and @tbl-rq2-en-smooth, respectively. The parametric term `interviewee_final_blp_score` indicates that there is no significant linear effect of language dominance, as measured by the Bilingual Language Profile (BLP), on A1-P0 values. The smooth term `ti(point_vwlpct):interviewee_final_blp_score` is similarly non-significant, indicating that there is no significant non-linear effect of language dominance on A1-P0 values.

```{r}
#| label: tbl-rq2-en-param
#| tbl-cap: Parametric terms for the model considering the impact of language dominance, as measured by the Bilingual Language Profile (BLP), on A1-P0 in English. The p-value for the varibale `interviewee_final_blp_score` above 0.05 indicates that the effect is non-significant.

rq2_en_param_ft <- flextable(gam_rq2_en_ar1_param_df) %>%
  colformat_double(digits = 2) %>%
  bold(part = "header") %>%
  autofit()

rq2_en_param_ft

```

```{r}
#| label: tbl-rq2-en-smooth
#| tbl-cap: Smooth terms for the model considering the impact of language dominance, as measured by the Bilingual Language Profile (BLP), on A1-P0 in English. The p-value for the varibale `ti(point_vwlpct):interviewee_final_blp_score` above 0.05 indicates that the effect is non-significant.

rq2_en_smooth_ft <- flextable(gam_rq2_en_ar1_smooth_df) %>%
  colformat_double(digits = 2) %>%
  bold(part = "header") %>%
  autofit()

rq2_en_smooth_ft

```

@fig-rq2-en-smooths plots three smooths for language dominance. These must be interpreted with caution, as neither the parametric nor the smooth effect for language dominance was found to be significant. Inspecting the plot, the three dominance levels converge at `r round(gam_rq2_en_ar1_dominance_effect_df_convergence_points[1,1],2)`% and `r round(gam_rq2_en_ar1_dominance_effect_df_convergence_points[2,1],2)`% of the vowel. In between those two points, A1-P0 estimates go from highest to lowest for English-dominant, balanced, and then Spanish-dominant bilinguals.

```{r}
#| label: fig-rq2-en-smooths
#| echo: false
#| fig-cap: Smooths representing A1-P0 values in English conditioned on language dominance at three levels. Spanish-dominant (BLP = -75), balanced (BLP = 0), and English-dominant (BLP = 75).

knitr::include_graphics(
  here("figs", "p_rq2_english_smooths.png")
  )

```

### Spanish

The model output's parametric and smooth terms are displayed in @tbl-rq2-es-param and @tbl-rq2-es-smooth, respectively. As was the case for the model fit to English data, the Spanish data similarly shows no significant effect for language dominance, as measured with the Bilingual Language Profile (BLP).

```{r}
#| label: tbl-rq2-es-param
#| tbl-cap: Parametric terms for the model considering the impact of language dominance, as measured by the Bilingual Language Profile (BLP), on A1-P0 in Spanish The p-value for the varibale `interviewee_final_blp_score` above 0.05 indicates that the effect is non-significant.

rq2_es_param_ft <- flextable(gam_rq2_es_ar1_param_df) %>%
  colformat_double(digits = 2) %>%
  bold(part = "header") %>%
  autofit()

rq2_es_param_ft

```

```{r}
#| label: tbl-rq2-es-smooth
#| tbl-cap: Smooth terms for the model considering the impact of language dominance, as measured by the Bilingual Language Profile (BLP), on A1-P0 in Spanish The p-value for the varibale `ti(point_vwlpct):interviewee_final_blp_score` above 0.05 indicates that the effect is non-significant.

rq2_es_smooth_ft <- flextable(gam_rq2_es_ar1_smooth_df) %>%
  colformat_double(digits = 2) %>%
  bold(part = "header") %>%
  autofit()

rq2_es_smooth_ft

```

@fig-rq2-es-smooths plots three smooths for language dominance, which against must be interpreted cautiously, as no significant effect for language dominance was found. Inspecting the plot, the three dominance levels converge at `r round(gam_rq2_es_ar1_dominance_effect_df_convergence_points[1,1],2)`% of the vowel. Before this point, Spanish-dominant bilinguals participants show the highest A1-P0 values, followed by balanced bilinguals, and then English-dominant bilinguals. After the convergence point, Spanish-dominant and English-dominant smooths switch positions.

```{r}
#| label: fig-rq2-es-smooths
#| echo: false
#| fig-cap: Smooths representing A1-P0 values in Spanish conditioned on language dominance at three levels. Spanish-dominant (BLP = -75), balanced (BLP = 0), and English-dominant (BLP = 75).

knitr::include_graphics(
  here("figs", "p_rq2_spanish_smooths.png")
  )

```

# Discussion

The present study combines articulatory phonology, sociolinguistics, and bilingualism to better understand whether or not language-specific gestural timing is stored in the bilingual brain, and if so, if language dominance impacts that storage. Specifically, coarticulatory vowel nasalization of /a/ was examined using the acoustic measure A1-P0 in simultaneous and early sequential bilinguals in the Multilingual Hispanic Speech in California [MuHSiC; @amengual2026muhsic] corpus.

The first research question was *Do bilinguals produce language-specific timing patterns of nasal gestures in English and Spanish?* Based on the statistical model, this can be answered affirmatively. Different timing patterns for nasal gestures were found for English and Spanish, in line with previous research [@beristain2023airflow]. This indicates that even though vowel nasalization is non-contrastive in Spanish and English, coarticulatory patterns are still stored in the bilingual phonological system.

The second research question was *Does language dominance predict degree of nasalization in each language in /aN/ sequences?* The statistical models indicate that language dominance does not predict degree of nasalization in either language. Although this is in line with previous empirical research [@beristain2023airflow], the opposite was hypothesized due to low variation in language dominance scores in the previous research, as well as theoretical motivation from the Complex Adaptive System Principles (CASP) model [@filipovic2019complex; @hawkins2024bilingualism]. The CASP model predicts that balanced bilinguals will show language-specific timing patterns, but that unbalanced bilinguals are more likely to [maximize common ground]{.smallcaps} between both of their languages. It was predicted, then, that non-contrastive articulatory timing, such as coarticulatory vowel nasalization, would be subject to convergence between the two languages for unbalanced bilinguals.

The findings regarding language dominance's lack of effect on coarticulatory nasalization timing patterns goes against the CASP model's predictions. It should be noted that the CASP model has not yet been explicitly tested on phonetic or phonological data, indicating that the model's predictions may not be appropriate for such data, as opposed to other domains of language.

## Limitations and Future Directions

The present study has a number of limitations, which are explored below. The Bilingual Language Profile (BLP) scores are skewed towards English-dominance, and the range is rather limited compared to the full possible range of scores.

Currently, the present study limits itself to productions of /p t k/ + /a/ + /m n/ in female-identifying bilinguals. The corpus includes male-identifying individuals, and of course the whole gamut of vowel-nasal /VN/ sequences in both English and Spanish. As such, future work should move to include the rest of the corpus.

Regarding the vowel choice, /a/ was chosen because it had the best performance for A1-P0, whereas it performs poorly for high vowels. This reduces the possibility of examining the full extent of the available data. Furthermore, A1-P0 remains fairly opaque as an acoustic correlate of the articulatory movement of lowering the velum. It does correlate with nasal airflow [@zellou2012similarity] and nasal consonant perception [@zellou2014gradient], but other aspects of speech production may affect it. Given such, it would be best to find another method, such as machine learning through `xgboost` [@chen2016xgboost], to incorporate a number of acoustic measures that correlate with nasality [@carignan2021practical].

An inherent difficulty with data from sociolinguistic interviews, and field data is general, is that the acoustic data noisier than laboratory data [@face2003intonation]. Not only may participants actually speak differently, but the recordings may be of lesser audio quality, making extraction of acoustic features less reliable. However, the amount of data obtained in a more-or-less natural environment is seen as a valuable compromise.

## Conclusion

The present study reported on the impact of language dominance on language-specific timing patterns of coarticulatory vowel nasalization, measured with A1-P0 over the time course of a vowel, in Spanish-English bilinguals in the Multilingual Hispanic Speech in California [MuHSiC; @amengual2026muhsic] corpus. Bilinguals were found to produce language-specific timing patterns, but language dominance not have a significant linear or non-linear effect on A1-P0 over the course of the vowel. This goes against predictions of the Complex Adaptive System Principles (CASP) model [@filipovic2019complex; @hawkins2024bilingualism], suggesting that the model may not accurately predict phonetic phenomena in bilinguals.

# References

::: {#refs}
:::